---
title: "PFAS_DataDownload"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#INTRODUCTION

This script pulls all California drinking water monitoring data for the previous 5 years from SDWIS, filters for PFAS data, and spatially joins drinking water well monitoring data with service area boundary layers.

#Resources

system boundaries homepage
https://gis.data.ca.gov/datasets/waterboards::california-drinking-water-system-area-boundaries?geometry=-148.883%2C31.064%2C-89.601%2C43.271

API for water system boundary layer
https://opendata.arcgis.com/datasets/fbba842bf134497c9d611ad506ec48cc_0.geojson

#Setup
```{r}
library(tidyverse)
library(foreign)
library(readxl)
library(vroom) #much faster than readr for csvs
library(sf) #read shapefiles
library(rgdal) #read OGR vector maps into spatial objects
library(magrittr)
```

# Data Import
## Chemical Data
```{r}
# download, upzip, and read most recent data and chemical storet info
urls <- c("https://www.waterboards.ca.gov/drinking_water/certlic/drinkingwater/documents/edtlibrary/chemical_as_csv.zip",
          "https://www.waterboards.ca.gov/drinking_water/certlic/drinkingwater/documents/edtlibrary/storet_as_dbf.zip",
          'https://www.waterboards.ca.gov/drinking_water/certlic/drinkingwater/documents/edtlibrary/siteloc_as_dbf.zip')
temp1 <- temp2 <- temp3 <- tempfile()

download.file(urls[1], temp1)
unzip(temp1, 
      exdir = "temp")

download.file(urls[2], temp2)
unzip(temp2,
      exdir = "temp")

download.file(urls[3], temp3)
unzip(temp3,
      exdir = "temp")

rm(temp1, temp2, temp3) # remove temp files

# read chem and storet data into R
# sometimes, R fails to unzip `chem`. unsure why, but manual download/unzip works
chem  <- vroom("temp/chemical.csv")
stor  <- read.dbf("temp/storet.dbf")
siteloc <- read.dbf("temp/siteloc.dbf")

# SDWIS data updates periodically, breaking the csv in url:
# https://data.ca.gov/dataset/drinking-water-public-water-system-information
sdwis <- vroom("https://data.ca.gov/dataset/d6d3beac-6735-4127-9324-4e70f61698d9/resource/9dca2f92-4630-4bee-a9f9-69d2085b57e3/download/drinking-water-watch-public-water-system-facilities.txt")

# make equivalent water system identifers 
sdwis$`Water System No` <- str_sub(sdwis$`Water System No`, 3, 9)
chem$PRIM_STA_C <- str_sub(chem$PRIM_STA_C, 1, 7)

# join chem and stor data
chem <- left_join(chem, stor, by = "STORE_NUM")
chem <- left_join(chem, sdwis, by = c("PRIM_STA_C" = "Water System No"))
chem <- chem %>% rename(chemical = CHEMICAL__)
# write the joined data (optional, takes a while)
#write_rds(chem, "chem.rds")
```

### Disply names of unique chemicals
```{r}
#get unique names of chemicals for filtering
unique(chem$chemical) %>% sort()
```

### Filter for PFAS
```{r}
#make list of PFAS and filter
pfaslist <- c('PERFLUOROBUTANESULFONIC ACID (PFBS)',
'PERFLUORODECANOIC ACID (PFDA)',
'PERFLUORODODECANOIC ACID (PFDoA)',
'PERFLUOROHEPTANOIC ACID (PFHpA)',
'PERFLUOROHEXANE SULFONIC ACID (PFHxS)',
'PERFLUOROHEXANOIC ACID (PFHxA)',
'PERFLUORONONANOIC ACID (PFNA)',
'PERFLUOROOCTANE SULFONIC ACID (PFOS)',
'PERFLUOROOCTANOIC ACID (PFOA)',
'PERFLUOROTETRADECANOIC ACID (PFTA)',
'PERFLUOROTRIDECANOIC ACID (PFTrDA)',
'PERFLUOROUNDECANOIC ACID (PFUnA)')

#filter for PFAS data and delete the rest
pfas <- chem %>% 
  filter(chemical %in% pfaslist)
rm(chem,sdwis, siteloc, stor, pfaslist, urls)

## Write PFAS data to csv ##
write.csv(pfas, "data/pfas.csv")
```
