---
title: "Drinking Water Download"
author: "Scott Coffin"
date: "`r Sys.Date()`"
output:   
  html_document:
    code_folding: hide
    theme: journal
    toc: yes
    toc_float: yes
    toc_depth: 6
    number_sections: true
    includes:
     # after_body: footer.html
  word_document:
    toc: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```
#Setup
## Library
```{r include=FALSE}
library(tidyverse)
library(RColorBrewer)
library(collapsibleTree)
library(foreign)
library(readxl)
library(vroom) #much faster than readr for csvs
library(sf) #read shapefiles
library(rgdal) #read OGR vector maps into spatial objects
library(magrittr)
```

This script pulls all California drinking water monitoring data for the previous 5 years from SDWIS, filters for PFAS data, and spatially joins drinking water well monitoring data with service area boundary layers.

##Resources

*Public Datasets*
system boundaries homepage
https://gis.data.ca.gov/datasets/waterboards::california-drinking-water-system-area-boundaries?geometry=-148.883%2C31.064%2C-89.601%2C43.271

API for water system boundary layer
https://opendata.arcgis.com/datasets/fbba842bf134497c9d611ad506ec48cc_0.geojson

https://www.waterboards.ca.gov/drinking_water/certlic/drinkingwater/documents/edtlibrary

*Private Datasets*
[NOTE: THESE DATASETS HAVE ALREADY BEEN DOWNLOADED INTO THIS REPO. THIS INFO IS FOR DOCUMENTATION PURPOSES ONLY]
Treatment data for water systems were obtained from mDWW SDWIS database.
To obtain this data, go to http://wb-sb-sdweb1:8080/CADWW/JSP/index.jsp and select "active treatment plants" from the report type, then download the excel spreadshee. Open in excel, then unmerge the cells and search and delete.

Lat/Long data for water systems has already been geocoded and is available internally:
http://reportmanager/Reports/report/DDW/DDW%20Reports/GIS/SDWIS_WSF_PSCode_Locational_Data

# Data Import
## Chemical Data
```{r}
# download, upzip, and read most recent data and chemical storet info
urls <- c("https://www.waterboards.ca.gov/drinking_water/certlic/drinkingwater/documents/edtlibrary/chemical_as_csv.zip",
          "https://www.waterboards.ca.gov/drinking_water/certlic/drinkingwater/documents/edtlibrary/storet_as_dbf.zip",
          'https://www.waterboards.ca.gov/drinking_water/certlic/drinkingwater/documents/edtlibrary/siteloc_as_dbf.zip')
temp1 <- temp2 <- temp3 <- tempfile()

download.file(urls[1], temp1)
unzip(temp1, 
      exdir = "temp")

download.file(urls[2], temp2)
unzip(temp2,
      exdir = "temp")

download.file(urls[3], temp3)
unzip(temp3,
      exdir = "temp")

rm(temp1, temp2, temp3) # remove temp files

# read chem and storet data into R
# sometimes, R fails to unzip `chem`. unsure why, but manual download/unzip works
chem  <- vroom("temp/chemical.csv")
stor  <- read.dbf("temp/storet.dbf")
siteloc <- read.dbf("temp/siteloc.dbf")

# SDWIS data updates periodically, breaking the csv in url:
# https://data.ca.gov/dataset/drinking-water-public-water-system-information
sdwis <- vroom("https://data.ca.gov/dataset/d6d3beac-6735-4127-9324-4e70f61698d9/resource/9dca2f92-4630-4bee-a9f9-69d2085b57e3/download/drinking-water-watch-public-water-system-facilities.txt")

# make equivalent water system identifers 
sdwis$`Water System No` <- str_sub(sdwis$`Water System No`, 3, 9)
chem$PRIM_STA_C <- str_sub(chem$PRIM_STA_C, 1, 7)

# join chem and stor data
chem <- left_join(chem, stor, by = "STORE_NUM")
chem <- left_join(chem, sdwis, by = c("PRIM_STA_C" = "Water System No"))
chem <- chem %>% rename(chemical = CHEMICAL__)
# write the joined data (optional, takes a while)
#write_rds(chem, "chem.rds")
```

### Disply names of unique chemicals
```{r}
#get unique names of chemicals for filtering
unique(chem$chemical) %>% sort()
```

### Filter for PFAS
```{r}
#make list of PFAS and filter
pfaslist <- c('PERFLUOROBUTANESULFONIC ACID (PFBS)',
'PERFLUORODECANOIC ACID (PFDA)',
'PERFLUORODODECANOIC ACID (PFDoA)',
'PERFLUOROHEPTANOIC ACID (PFHpA)',
'PERFLUOROHEXANE SULFONIC ACID (PFHxS)',
'PERFLUOROHEXANOIC ACID (PFHxA)',
'PERFLUORONONANOIC ACID (PFNA)',
'PERFLUOROOCTANE SULFONIC ACID (PFOS)',
'PERFLUOROOCTANOIC ACID (PFOA)',
'PERFLUOROTETRADECANOIC ACID (PFTA)',
'PERFLUOROTRIDECANOIC ACID (PFTrDA)',
'PERFLUOROUNDECANOIC ACID (PFUnA)')

#filter for PFAS data and delete the rest
pfas <- chem %>% 
  filter(chemical %in% pfaslist)
rm(chem,sdwis, siteloc, stor, pfaslist, urls)

## Write PFAS data to csv ##
write.csv(pfas, "output/pfas.csv")
```

#Treatment data
Treatment data for water systems were obtained from mDWW SDWIS database.
To obtain this data, go to http://wb-sb-sdweb1:8080/CADWW/JSP/index.jsp and select "active treatment plants" from the report type, then download the excel spreadshee. Open in excel, then unmerge the cells and search and delete.
```{r}
#import data
treatment <- read.csv("data/treatment_tech_info.csv",
                      stringsAsFactors = TRUE,
                      na.strings = c("NULL", "NA"))
##Note that this is a *nested* dataset, as many treatment plants employ numerous unit processes##

#nest data
nested_treatment <- treatment %>% 
   select(c(Principal.County.Served, pwsid, Water.System.Name, State.Water.System.Type,
             Primary.Water.Source.Type, Total.Number.of.Service.Connections, Total.Population,
             CITY, facility_names, System.Status, Fee.Code, 
            #variables to nest
            unit_process, treatment_objective, treatment_process)) %>% 
  group_by(Principal.County.Served, pwsid, Water.System.Name, State.Water.System.Type,
             Primary.Water.Source.Type, Total.Number.of.Service.Connections, Total.Population,
             CITY, System.Status, Fee.Code
           ) %>% 
  nest(.key = "treatment")
```


## Classification
Based on what is known regarding PFAS removal capabilities in the literature, the following treatment processes are regarded as having a high  likelihood of significantly removing PFAS:

REVERSE OSMOSIS ,
REVERSE OSMOSIS,POINT OF ENTRY ,
REVERSE OSMOSIS,POINT OF USE ,
ACTIVATED CARBON, GRANULAR ,
ION EXCHANGE 

To assess the approximate proportion of systems in the state that have some capability of removing PFAS, systems with these treatments were summarized.

References for treatment data: *NEEDED*

Removal percentages are coded as follows:
* High = >90%
* Medium = 30 to 90%
* Low = 0 to 30%

```{r}
likely.treatment <- treatment %>% 
  mutate(categorical_removal = case_when(
                                        treatment_process== "REVERSE OSMOSIS" ~ "high",
                                        treatment_process== "REVERSE OSMOSIS,POINT OF ENTRY" ~ "high",
                                        treatment_process== "REVERSE OSMOSIS,POINT OF USE" ~ "high",
                                        treatment_process== "ION EXCHANGE" ~ "high",
                                        treatment_process== "ACTIVATED CARBON, GRANULAR" ~ "medium",
                                        treatment_process== "ACTIVATED CARBON, POWDERED" ~ "medium",
                                        )) %>% 
  replace_na(list(categorical_removal = "low"))
```

## CollapsibleTree
```{r}
treatment.mod <- likely.treatment %>% 
  filter(Total.Number.of.Service.Connections > 60000) %>% 
  drop_na(unit_process)

collapsible.tree <- collapsibleTree(treatment.mod, 
                                    c("wool", "tension", "breaks"),
                root = "Treatment Heirarchy", 
                hierarchy = c("treatment_objective", 
                              "unit_process",
                              "treatment_process", 
                              "categorical_removal"
                              ),
                 fontSize = 9,
                zoomable = TRUE,
                #tooltip = TRUE,
                #fill = brewer.pal("RdBu"),
                fillByLevel = FALSE) 

#export for easy viewing elsewhere
htmltools::save_html(collapsible.tree, file='output/collapsibleTree.html')
htmlwidgets::saveWidget(collapsible.tree, file="output/collapsibleTree_widget.html")

## visualize here
collapsible.tree
```
