---
title: "PFAS Geosptial analysis"
author: "Scott Coffin"
date: "`r Sys.Date()`"
output:   
  html_document:
    code_folding: hide
    theme: journal
    toc: yes
    toc_float: yes
    toc_depth: 6
    number_sections: true
    includes:
     # after_body: footer.html
  word_document:
    toc: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```
#Setup
```{r include=FALSE}
library(tidyverse)
library(DT)
library(RColorBrewer)
library(collapsibleTree)
library(janitor)
library(ggpmisc)
library(scales)
library(tidygeocoder)
library(mapview)      # interactive maps!
library(ggspatial)
library(basemaps)
library(USAboundaries)
library(sf)
library(ggmap)
library(ggpubr)
library(ggsci)
```

# Read in Data
# Drinking water chemical data
```{r}
pfas <- read.csv("output/pfas.csv")
```
 
## Spatial Data

system boundaries homepage
https://gis.data.ca.gov/datasets/waterboards::california-drinking-water-system-area-boundaries?geometry=-148.883%2C31.064%2C-89.601%2C43.271

API for water system boundary layer
https://opendata.arcgis.com/datasets/fbba842bf134497c9d611ad506ec48cc_0.geojson

### System Area Boundary Layer

```{r eval=FALSE, include=FALSE}
### This API is no longer active, so need to pull manually ###
SABL_link <- "https://opendata.arcgis.com/datasets/fbba842bf134497c9d611ad506ec48cc_0.zip"
temp <- tempfile()
temp2 <- tempfile()
download.file(SABL_link, temp)
unzip(zipfile = temp, exdir = temp2)
#open with rgdal
SABL <- readOGR(dsn = temp2, layer = "California_Drinking_Water_System_Area_Boundaries")
#or open with sf
SABL <- sf::st_read(paste0(temp2, "/California_Drinking_Water_System_Area_Boundaries.shp"))
unlink(temp)
unlink(temp2)
```

data should be pulled manually by emailing request to: ddw-dsu@waterboards.ca.gov

DISCLAIMER:
Please be advised that this copy of the System Area Boundary Layer data is a snapshot of the CA drinking water system area boundaries provided to the State Board by individual water systems. A water system may have provided accurate, up-to-date boundaries, or may have provided boundaries that are only approximate. This layer does not contain all Public Water System boundaries. The dataset is in the process of being reviewed and verified. In addition, new boundaries are being added. Please note that these definitions are NOT a legal definition and should NOT be used to settle boundary disputes. Because the Water Board is targeting different information, there is no conflict with other authoritative agencies' definitions.

```{r}
#open with rgdal
SABL <- readOGR(dsn = "data/GIS", layer = "SABL_Public_220802") #destination with zip extract of system area boundary layer from DDW

#or open with sf
SABL_sf <- sf::st_read("data/GIS/SABL_Public_220802.shp")
```
### Drinking Water Well Spatial Locations

NOTE: The following section is outdated, and is kept here just for reference purposes.

<!-- # ```{r eval=FALSE, include=FALSE} -->
<!-- # ### Exact and up-to-date location data is avilable internally -->
<!-- # #http://reportmanager/Reports/report/DDW/DDW%20Reports/GIS/SDWIS_WSF_PSCode_Locational_Data -->
<!-- # #static download file in folder -->
<!-- # locationData <- read.csv("data/SDWIS_WSF_PSCode_Locational_Data.csv") %>% -->
<!-- #   rename("pwsid" = "WS_ID", -->
<!-- #          "latitude" = "LATITUDE_MEASURE", -->
<!-- #          "longitude" = "LONGITUDE_MEASURE") %>% -->
<!-- #   dplyr::select(c(pwsid, latitude, longitude)) %>% -->
<!-- #   drop_na(latitude) -->
<!-- # -->
<!-- # ## PS Code = PRIM_STA_C = Primary Station Code ## -->
<!-- # ## Concat "CA" with PRIM_STA_C in pfas data to join with geospatial data ## -->
<!-- # pfas %<>% -->
<!-- #   mutate(pwsid = paste0("CA", PRIM_STA_C)) -->
<!-- # -->
<!-- # #join public water system ID location data with PFAS monitoring data from wells -->
<!-- # pfas_loc <- plyr::join(pfas, locationData, -->
<!-- #                                    by = "pwsid", type = "left", match = "first") -->

<!-- #transform data frame with well points and water quality data into sf object -->
<!-- pfas_sf <- st_as_sf(pfas_loc %>% drop_na(c(latitude, longitude)), #remove missing lat/long data -->
<!--                       coords = c('longitude', 'latitude'), crs = 4326) -->

<!-- #intersect well points with SPA polygons -->
<!-- wells_int <- st_join(pfas_sf, left = TRUE, SABL_wgs["SPA_Name"]) -->
<!-- #the wells_int file should then replicate the data within "PFAS_SPAspatialjoin.dbf" -->
<!-- # ``` -->


### Spatial Join of SABL and Well Data
```{r}
#transform shapefile (if necessary)
SABL_wgs <- st_transform(SABL_sf, crs = 4326) %>%  #WGS85 projection
  rename("pwsid" = "SABL_PWSID") #rename to be consistent with pfas data

#join drinking water well data to service area boundary layer based on public water system ID
pfas_SABL <- left_join(pfas, SABL_wgs, by = "pwsid")
```

### Average well concentrations within an SPA

NOTE: Many wells are below *laboratory detection limits* which vary by chemical and lab. Ordinarily we would know if it's below the detection limit for reporting (DLR), however PFAS lack DLR right now. Need to figure out data censoring method.
```{r}
SPA_avg <- pfas_SABL %>% 
    group_by(pwsid, chemical) %>%
  summarize(spa_chem_mean = mean(FINDING))

head(SPA_avg)
```
```{r}
sp <- ggplot(SPA_avg, aes(x = spa_chem_mean, y = spa_chem_mean)) +
  geom_point() +
  facet_wrap(~chemical)

sp
```


[NOTE: THE FOLLOWING CODE IS EXAMPLE CODE COPIED FROM A DIFFERENT REPO. PLEASE USE WHAT'S NECESSARY/INSPIRING AND DELETE]

#Spatial Analysis
## Data preparation
Plot locations based on water system location (not exact sampling location)

```{r eval=FALSE, include=FALSE}
#read in watsys location data from https://www.waterboards.ca.gov/drinking_water/certlic/drinkingwater/EDTlibrary.html
watsys <- read.csv("data/watsys.csv") %>% 
  #rename("pwsid" = "ï..pwsid") %>% 
  dplyr::select(c(pwsid, ADDRESS, STATE, ZIP, ZIP_EXT))

#join with final list
final.list.locations <- left_join(final.list, watsys, by = "pwsid")

#geocode lat and long based on addresses of water systems (WARNING: This chunk takes a long time to run!)
final.list.locations <- final.list.locations %>% 
  #paste street address with city, state and zip code
  mutate(newAddress = paste(ADDRESS, CITY, STATE, ZIP)) %>% 
  tidygeocoder::geocode(newAddress, method = 'osm', lat = latitude, long = longitude) # open-source geocoding service https://nominatim.org/
```

ALTERNATIVE METHOD: water systems have already been geocoded internally.
```{r}
### Exact and up-to-date location data is avilable internally
#http://reportmanager/Reports/report/DDW/DDW%20Reports/GIS/SDWIS_WSF_PSCode_Locational_Data
#static download file in folder
locationData <- read.csv("data/SDWIS_WSF_PSCode_Locational_Data.csv") %>% 
  rename("pwsid" = "WS_ID",
         "latitude" = "LATITUDE_MEASURE",
         "longitude" = "LONGITUDE_MEASURE") %>% 
  dplyr::select(c(pwsid, latitude, longitude)) %>% 
  drop_na(latitude)
  

#join with final list
#if we join as-is, there will be duplicate matches, so we need to simplify
final.list.locations <- plyr::join(final.list, locationData,
                                   by = "pwsid", type = "left", match = "first")
  #left_join(final.list, locationData, by = "pwsid") %>% 
  #dplyr::distinct("pwsid",
   #               .keep_all = TRUE)
```


### Static maps
#### Simple
```{r}
library(ggthemes)
points <- final.list.locations %>% 
  drop_na(c(longitude, latitude)) %>% 
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326)

#get basemap for CA
states <- map_data("state")
ca_df <- subset(states, region == "california")
counties <- map_data("county")
ca_county <- subset(counties, region == "california")

ggplot() + 
  #basemap
  geom_polygon(data = ca_df, mapping = aes(x = long, y = lat, group = group),
               color = "black", fill = "gray") +
  coord_fixed(1.3) +
# counties
 # geom_polygon(fill = NA, color = "white", 
  #             data = ca_county, mapping = aes(x = long, y = lat)) +
  #geom_polygon(color = "black", fill = NA, alpha = 1,
   #            data = ca_county, mapping = aes(x = long, y = lat)) +  # get the state border back on top
  #add points
  geom_point(data = final.list.locations,
             aes(x = longitude, y = latitude),
             color = "blue", size = 1.5) +
  theme_minimal()

```
#### Annotated Maps
##### Population
```{r}
population <- ggplot() + 
  #basemap
  geom_polygon(data = ca_df, mapping = aes(x = long, y = lat, group = group),
               color = "black", fill = "gray50") +
  coord_fixed(1.3) +
  geom_point(data = final.list.locations,
             aes(x = longitude, y = latitude, color = log10(Total.Population)),
              size = 2,
             alpha = 0.9) +
  scale_color_gsea(name = "Log10(Population Served)") +
  #scale_color_manual(values = cal_palette("sierra1", n = 50, type = "continuous")) +
  ## Pretty annotations
  annotation_scale(location = "bl") + #add scale
  annotation_north_arrow(location = "bl", #north arrow
                         pad_x =  unit(0.5, "in"), pad_y = unit(0.5, "in"), #shift it so it doesn't plot on top of scale bar
                         style = north_arrow_fancy_orienteering()) + #change style of north arrow
  #labs(legend = "Threshold") +
  xlab("Longitude (NAD83)") +
  ylab("Latitude (NAD83)") +
  theme_minimal() +
  theme(
        legend.background = element_rect(fill = "white", color = "black"),
        legend.position = c(0.81, 0.7),
        legend.key.size = unit(0.3, 'cm')#,
        #legend.title = element_blank()
        )

population
```