---
title: "PFAS_DataDownload"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#INTRODUCTION

This script pulls all California drinking water monitoring data for the previous 5 years from SDWIS, filters for PFAS data, and spatially joins drinking water well monitoring data with service area boundary layers.

#Resources

system boundaries homepage
https://gis.data.ca.gov/datasets/waterboards::california-drinking-water-system-area-boundaries?geometry=-148.883%2C31.064%2C-89.601%2C43.271

API for water system boundary layer
https://opendata.arcgis.com/datasets/fbba842bf134497c9d611ad506ec48cc_0.geojson

#Setup
```{r}
library(tidyverse)
library(foreign)
library(readxl)
library(vroom) #much faster than readr for csvs
library(sf) #read shapefiles
library(rgdal) #read OGR vector maps into spatial objects
library(magrittr)
```

# Data Import
## Chemical Data
```{r}
# download, upzip, and read most recent data and chemical storet info
urls <- c("https://www.waterboards.ca.gov/drinking_water/certlic/drinkingwater/documents/edtlibrary/chemical_as_csv.zip",
          "https://www.waterboards.ca.gov/drinking_water/certlic/drinkingwater/documents/edtlibrary/storet_as_dbf.zip",
          'https://www.waterboards.ca.gov/drinking_water/certlic/drinkingwater/documents/edtlibrary/siteloc_as_dbf.zip')
temp1 <- temp2 <- temp3 <- tempfile()

download.file(urls[1], temp1)
unzip(temp1, 
      exdir = "temp")

download.file(urls[2], temp2)
unzip(temp2,
      exdir = "temp")

download.file(urls[3], temp3)
unzip(temp3,
      exdir = "temp")

rm(temp1, temp2, temp3) # remove temp files

# read chem and storet data into R
# sometimes, R fails to unzip `chem`. unsure why, but manual download/unzip works
chem  <- vroom("temp/chemical.csv")
stor  <- read.dbf("temp/storet.dbf")
siteloc <- read.dbf("temp/siteloc.dbf")

# SDWIS data updates periodically, breaking the csv in url:
# https://data.ca.gov/dataset/drinking-water-public-water-system-information
sdwis <- vroom("https://data.ca.gov/dataset/d6d3beac-6735-4127-9324-4e70f61698d9/resource/9dca2f92-4630-4bee-a9f9-69d2085b57e3/download/drinking-water-watch-public-water-system-facilities.txt")

# make equivalent water system identifers 
sdwis$`Water System No` <- str_sub(sdwis$`Water System No`, 3, 9)
chem$PRIM_STA_C <- str_sub(chem$PRIM_STA_C, 1, 7)

# join chem and stor data
chem <- left_join(chem, stor, by = "STORE_NUM")
chem <- left_join(chem, sdwis, by = c("PRIM_STA_C" = "Water System No"))
chem <- chem %>% rename(chemical = CHEMICAL__)
# write the joined data (optional, takes a while)
#write_rds(chem, "chem.rds")
```

### Disply names of unique chemicals
```{r}
#get unique names of chemicals for filtering
unique(chem$chemical) %>% sort()
```

### Filter for PFAS
```{r}
#make list of PFAS and filter
pfaslist <- c('PERFLUOROBUTANESULFONIC ACID (PFBS)',
'PERFLUORODECANOIC ACID (PFDA)',
'PERFLUORODODECANOIC ACID (PFDoA)',
'PERFLUOROHEPTANOIC ACID (PFHpA)',
'PERFLUOROHEXANE SULFONIC ACID (PFHxS)',
'PERFLUOROHEXANOIC ACID (PFHxA)',
'PERFLUORONONANOIC ACID (PFNA)',
'PERFLUOROOCTANE SULFONIC ACID (PFOS)',
'PERFLUOROOCTANOIC ACID (PFOA)',
'PERFLUOROTETRADECANOIC ACID (PFTA)',
'PERFLUOROTRIDECANOIC ACID (PFTrDA)',
'PERFLUOROUNDECANOIC ACID (PFUnA)')

#filter for PFAS data and delete the rest
pfas <- chem %>% 
  filter(chemical %in% pfaslist)
rm(chem,sdwis, siteloc, stor, pfaslist, urls)

## Write PFAS data to csv ##
write.csv(pfas, "data/pfas.csv")
```

## Spatial Data

system boundaries homepage
https://gis.data.ca.gov/datasets/waterboards::california-drinking-water-system-area-boundaries?geometry=-148.883%2C31.064%2C-89.601%2C43.271

API for water system boundary layer
https://opendata.arcgis.com/datasets/fbba842bf134497c9d611ad506ec48cc_0.geojson

### System Area Boundary Layer

```{r eval=FALSE, include=FALSE}
### This API is no longer active, so need to pull manually ###
SABL_link <- "https://opendata.arcgis.com/datasets/fbba842bf134497c9d611ad506ec48cc_0.zip"
temp <- tempfile()
temp2 <- tempfile()
download.file(SABL_link, temp)
unzip(zipfile = temp, exdir = temp2)
#open with rgdal
SABL <- readOGR(dsn = temp2, layer = "California_Drinking_Water_System_Area_Boundaries")
#or open with sf
SABL <- sf::st_read(paste0(temp2, "/California_Drinking_Water_System_Area_Boundaries.shp"))
unlink(temp)
unlink(temp2)
```

data should be pulled manually by emailing request to: ddw-dsu@waterboards.ca.gov

DISCLAIMER:
Please be advised that this copy of the System Area Boundary Layer data is a snapshot of the CA drinking water system area boundaries provided to the State Board by individual water systems. A water system may have provided accurate, up-to-date boundaries, or may have provided boundaries that are only approximate. This layer does not contain all Public Water System boundaries. The dataset is in the process of being reviewed and verified. In addition, new boundaries are being added. Please note that these definitions are NOT a legal definition and should NOT be used to settle boundary disputes. Because the Water Board is targeting different information, there is no conflict with other authoritative agencies' definitions.

```{r}
#open with rgdal
SABL <- readOGR(dsn = "data/GIS", layer = "SABL_Public_220802") #destination with zip extract of system area boundary layer from DDW

#or open with sf
SABL_sf <- sf::st_read("data/GIS/SABL_Public_220802.shp")
```
### Drinking Water Well Spatial Locations

NOTE: The following section is outdated, and is kept here just for reference purposes.

<!-- # ```{r eval=FALSE, include=FALSE} -->
<!-- # ### Exact and up-to-date location data is avilable internally -->
<!-- # #http://reportmanager/Reports/report/DDW/DDW%20Reports/GIS/SDWIS_WSF_PSCode_Locational_Data -->
<!-- # #static download file in folder -->
<!-- # locationData <- read.csv("data/SDWIS_WSF_PSCode_Locational_Data.csv") %>% -->
<!-- #   rename("pwsid" = "WS_ID", -->
<!-- #          "latitude" = "LATITUDE_MEASURE", -->
<!-- #          "longitude" = "LONGITUDE_MEASURE") %>% -->
<!-- #   dplyr::select(c(pwsid, latitude, longitude)) %>% -->
<!-- #   drop_na(latitude) -->
<!-- # -->
<!-- # ## PS Code = PRIM_STA_C = Primary Station Code ## -->
<!-- # ## Concat "CA" with PRIM_STA_C in pfas data to join with geospatial data ## -->
<!-- # pfas %<>% -->
<!-- #   mutate(pwsid = paste0("CA", PRIM_STA_C)) -->
<!-- # -->
<!-- # #join public water system ID location data with PFAS monitoring data from wells -->
<!-- # pfas_loc <- plyr::join(pfas, locationData, -->
<!-- #                                    by = "pwsid", type = "left", match = "first") -->

<!-- #transform data frame with well points and water quality data into sf object -->
<!-- pfas_sf <- st_as_sf(pfas_loc %>% drop_na(c(latitude, longitude)), #remove missing lat/long data -->
<!--                       coords = c('longitude', 'latitude'), crs = 4326) -->

<!-- #intersect well points with SPA polygons -->
<!-- wells_int <- st_join(pfas_sf, left = TRUE, SABL_wgs["SPA_Name"]) -->
<!-- #the wells_int file should then replicate the data within "PFAS_SPAspatialjoin.dbf" -->
<!-- # ``` -->


### Spatial Join of SABL and Well Data
```{r}
#transform shapefile (if necessary)
SABL_wgs <- st_transform(SABL_sf, crs = 4326) %>%  #WGS85 projection
  rename("pwsid" = "SABL_PWSID") #rename to be consistent with pfas data

#join drinking water well data to service area boundary layer based on public water system ID
pfas_SABL <- left_join(pfas, SABL_wgs, by = "pwsid")
```

### Average well concentrations within an SPA

NOTE: Many wells are below *laboratory detection limits* which vary by chemical and lab. Ordinarily we would know if it's below the detection limit for reporting (DLR), however PFAS lack DLR right now. Need to figure out data censoring method.
```{r}
SPA_avg <- pfas_SABL %>% 
    group_by(pwsid, chemical) %>%
  summarize(spa_chem_mean = mean(FINDING))

head(SPA_avg)
```
```{r}
sp <- ggplot(SPA_avg, aes(x = spa_chem_mean, y = spa_chem_mean)) +
  geom_point() +
  facet_wrap(~chemical)

sp
```

